---
title: "Impact of Vitamin C Delivery Method on Tooth Growth"
author: "Jim Callahan"
date: "October 1, 2015"
output: pdf_document
---

### Overview
The data in this project is an R supplied data set, "Tooth Growth".
The official description in the R version 3.2.2 help file is:

> The response is the length of odontoblasts (teeth) in each of 10 guinea pigs 
> at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of 
> two delivery methods (orange juice or ascorbic acid).

### Data
The data in this project is an R supplied data set, "Tooth Growth" which can be 
accessed from the "datasets" package using the data command:

```{r}
# Access the "datasets" package using the R library command
# and then the data set itself using the R data command.
library(datasets)
data(ToothGrowth)

# Exploratory Data Analysis (EDA)
# Describe the structure of the data set, the first few rows
str(ToothGrowth)
head(ToothGrowth)
summary(ToothGrowth)
# Note the help file for the R data set can be accessed by typing
# help(ToothGrowth) 
# at the R command line.
```

The **R** help file also has sample code to produce a graph.
```{r}
# Graph Code from R help file
require(graphics)
coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth,
       xlab = "ToothGrowth data: length vs dose, given type of supplement")
```

The obvious thing to do here is to run a regression with:  

        dose = f(supp, data)  

or an ANOVA, but the instructions are explicit:
> Only use the techniques from class, even if there's 
> other approaches worth considering.

From the co-plot we can see that OJ appears to be clearly superior at the ".5 mg" and
"1.0 mg" dose levels, while at the "2.0 mg" dose level the means appear to be closer,
but even at 2.0 the Vitamin C observations seem to be more spread out suggesting
a wider variance. 

The questions then are:
1. How large are the differences in means?
2. Are the differences of means statistically significant?

```{r}
# Create a new variable "dosefactor" coercing dose to a factor
ToothGrowth$dosefactor <- as.factor(ToothGrowth$dose)
# Create a table showing how many observations we have at each level
# of supp and dosefactor
     ToothCount <- table(ToothGrowth$supp, ToothGrowth$dosefactor)
     ToothCount

```
Although we could compute a mean for all of the "VC" observations
and another mean for all of the "OJ" observations and compare the two, the
procedure is suspect since we know there were different treatment levels
and from the graph alone we can see there was likely some impact from
the treatment levels so these are not "identically distributed" as we would
want for an IID assumption.

So, we really need to compute six means, one for each combination of the level.
And then perform three tests, one for each level of dose comparing "OJ" and "VC".

```{r}
LengthOJat05 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dosefactor == "0.5"]
LengthOJat10 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dosefactor == "1"]
LengthOJat20 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dosefactor == "2"]

LengthVCat05 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dosefactor == "0.5"]
LengthVCat10 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dosefactor == "1"]
LengthVCat20 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dosefactor == "2"]

t.test(LengthOJat05, LengthVCat05)
t.test(LengthOJat10, LengthVCat10)
t.test(LengthOJat20, LengthVCat20)

```

The t-tests confirm there is a statistically significant difference in the means
at the "0.5 mg" and "1.0 mg" levels, but the the significance disappears at the
"2.0 mg" level. We can build a table to summarize the results.

```{r}
tdiff05 <- t.test(LengthOJat05, LengthVCat05)$statistic
tdiff10 <- t.test(LengthOJat10, LengthVCat10)$statistic
tdiff20 <- t.test(LengthOJat20, LengthVCat20)$statistic

pdiff05 <- t.test(LengthOJat05, LengthVCat05)$p.value
pdiff10 <- t.test(LengthOJat10, LengthVCat10)$p.value
pdiff20 <- t.test(LengthOJat20, LengthVCat20)$p.value

ttest  <- c(tdiff05, tdiff10, tdiff20)
pvalue <- c(pdiff05, pdiff10, pdiff20)
interpretation <- ifelse(pvalue <= .05, "significant", "not significant")

results <- cbind(ttest, pvalue, interpretation)
rownames(results) <- c("0.5 mg", "1.0 mg", "2.0 mg")
results
```

We have chosen to claim "significance" for pvalues less than .05 ("pvalue <= .05") 
or an alpha level of 5%.

### ASSUMPTIONS  
This is assuming that the variables are IID (Independent and Identially Distributed)
and follow a normal ("Gaussian") distribution that can be approximated by a Student's T
distribution.
  
 
